---
title: "MetMSLine Basics"
author: "WMB Edmands"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetMSLine Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following illustrates the automated MetMSLine workflow with example data
for those less familiar with R and the command line interface.
R is primarily an interactive programming language, however a degree of automation
can be acheived by the usage of wrapper functions, that is software functions
that 'wrap' together several other software functions. 

## Getting started. 

Install all of the latest package dependencies needed for proper/ complete 
functionality of MetMSLine. Copy and paste the following command into R and hit enter:
```{r, eval=F}
install.packages(c( 'ff', 'dynamicTreeCut', 'data.table', 'reshape2', 'foreach'))
```

This may take several minutes.

Read in example peak table and co-variates and pre-process the data. Run the
following code in the same R session at the command line (i.e. copy and paste the lines and hit enter):
```{r, include=F}
library(MetMSLine)
```
```{r, collapse=TRUE}
# read in table of synthetic MS1 metabolomic profiling data
peakTable <- system.file("extdata", "synthetic_MS_data.csv", package = "MetMSLine")
peakTable <- read.csv(peakTable, header=T, stringsAsFactors=F)


# load synthetic co-variates table in comma delimited csv file
coVariates <- system.file("extdata", "synthetic_coVar_data.csv", package = "MetMSLine")
coVariates <- read.csv(coVariates, header=T)
```

Next we need to establish the names of the samples, these observation names
will be used to identify the correct column names in the peak table.

By entering the command colnames we can see all of the names of the columns in
our peak table:
```{r}
colnames(peakTable)
```

The synthetic peak table appears very similar to output from the XCMS peak picking
software (specifically the **?xcms::diffreport** function).  

We can see that there are many columns containing important information about each
LC-MS variable such as the median mass-to-charge ("mzmed") and the median retention time
("rtmed") for example. Regardless of peak picking software (xcms or otherwise) a peak-picker 
output table in its most basic interpretation will almost always consist of rows 
of LC-MS variables and columns corresponding LC-MS variable information and 
sample peak intensity/ height information.

In order for MetMSLine to work it must know the column names which correspond to
the observation/ sample peak intensity information. In certain cases such as 
lowess smoothing MetMSLine must also be supplied with the names of quality 
control samples also in order to distinguish these from other samples.

We can see that from column 18 to the last column there are observation columns.
Within which we can see quality controls containing the character string "QC_", 
blanks (i.e. negative controls) containing the character string "blank_" and also
samples containing the character string "sample_". These unique character strings
can be then be used to quickly identify the correct column names to supply to MetMSLine
within your R session, like so:

```{r}
# observation names (i.e. sample names) by regular expression (?grep)
# all observation names
obsNames <- colnames(peakTable)[grep("QC_|sample_|blank_", colnames(peakTable))]
# print all the observation names
obsNames
```

We can also identify the quality control samples and assign them to a new
variable qcNames 
```{r}
qcNames <- colnames(peakTable)[grep("QC_", colnames(peakTable))] 
# print the QC names
qcNames
```

Then the sample names (assigned to sampNames)
```{r}
sampNames <- colnames(peakTable)[grep("sample_", colnames(peakTable))]
# print the sample names
sampNames
```

Then finally the blank sample names (assigned to blankNames).
```{r}
blankNames <- colnames(peakTable)[grep("blank_", colnames(peakTable))]
# print the blank names
blankNames
```


Now that we have identified our observation column names we can now start the
MetMSLine process by supplying the necessary arguments to the wrapper functions.

## 1. LC-MS peak table preprocessing.
```{r}
# detect number of cores using parallel package
nCores <- parallel::detectCores()
# conduct LC-MS data preprocessing
preProc_peakTable <- preProc(peakTable, obsNames, sampNames, qcNames, blankNames,
                             cvThresh=30, nCores=nCores)
```

The table returned from the preProc function contains three additional columns:
```{r}
tail(colnames(preProc_peakTable), n=3)
```

1. smoothSpanLoessFit: column contains the optimal loess smooth span parameter
identified by 7-fold cross validation for each LC-MS variable.

2. meanFCsampBlank: the fold change value of samples:blanks.

3. coeffVar: the coefficient of variation (CV%) of the repeat pooled QC injections for each LC-MS variable.

Furthermore, the original peakTable has also been subset and LC-MS variables
which were below the sample:blank fold change threshold (the default is >2 fold) 
or above the CV% threshold set (the default is 30%).

if you want to find out more about this wrapper function you can simply type ?MetMSLine::preProc
in your R console and hit enter to see the help page.

## 2. PCA based automated outlier detection and cluster identification.
